---
title: "Matching genetic variants using parquet files and duckdb"
author: "Thomas Sandmann"
date: "2025-08-17"
execute:
  echo: true
  warning: false
  message: false
  cache: false
freeze: true
categories: [R, TIL]
editor:
  markdown:
    wrap: 72
format:
  html:
    anchor-sections: true
    toc: true
    toc-depth: 4
    code-tools:
      source: true
      toggle: false
      caption: none
---

[ClinVar](https://www.ncbi.nlm.nih.gov/clinvar/intro/)
is a public archive of reports of human variations associated with 
diseases and drug responses, with supporting evidence, hosted at the
NCBI.

## Retrieving ClinVar variants from the NCBI's FTP server

First, we read information about disease-associated variants, available in
a gzip-compressed tab-delimited text file 
[from the NCBI](https://ftp.ncbi.nlm.nih.gov/pub/clinvar/tab_delimited/), 
and store information about single-nucleotide variants (SNVs) in a local
parquet file.

::: {.callout-note collapse="true"}
### VCF format

Frequently, information about genetic variation is shared using files in
[Variant Call Format (VCF)](https://en.wikipedia.org/wiki/Variant_Call_Format)
or variations of it, e.g. genomic VCF (gVCF).

While widely used, VCF files for large collections of variants can become
unwieldly. In 2021, researcher from 23&me explored other file formats and
reported that storing variant data in
[parquet files](https://medium.com/23andme-engineering/genetic-datastore-4b213256db31)
might be adventageous.

The NCBI makes ClinVar's information available in multiple formats, including
VCF and tab-delimited text files. Here, I am retrieving the latter. Multiple
solutions to convert VCF to character-delimited or parquet files exist, 
including methods in the
[GLOW Spark framework](https://glow.readthedocs.io/en/latest/etl/variant-data.html),
or the `VariantsToTable` tool included in the 
[GATK suite](https://gatk.broadinstitute.org/hc/en-us/articles/30332008315291-VariantsToTable).
The
[vcf-to-duckdb python module](https://pypi.org/project/vcf-to-duckdb/)
goes one step further and converts a VCF file to a DuckDB database, encompassing
both as Parquet files and accompanying SQL schema.

:::


```{r}
library(DBI)
library(duckdb)
library(glue)
```

Let's start by creating an in-memory `duckdb` instance and install the `httpfs`
extension to retrieve remote files.

```{r}
con <- dbConnect(duckdb())
DBI::dbExecute(con, "INSTALL httpfs")
```

The latest release of ClinVar is available from its website. For this tutorial,
I will convert it into a local parquet file stored in a temporary directory[^1].

[^1]: For real use cases, you probably want to store the parquet file in a 
permanent location, either locally or in a cloud bucket.

```{r}
kTab <- paste0(
	"https://ftp.ncbi.nlm.nih.gov/pub/clinvar/tab_delimited/",
	"variant_summary.txt.gz"
)
kParquet <- tempfile(fileext = ".parquet")
```

ClinVar accepts reports about any type of genetic variant, including 
single nucleotide variants (SNVs), small insertions/deletions or large copy 
number variations for human genome assemblies GRCh37 and GRCh38. 
Here, I will only retain rows describing single nucleotide variants and their
location in the GRCh38 assembly.

A single SQL query retrieves the ~ 380 MB tab-delimited file, filters for 
single-nucleotide variants on the GRCh38 assembly, and writes them into a 
~ 325 MB parquet file in a single step  [^2]

[^2]: The same command could be used to convert very large files, as `duckdb` 
does not need to retain all of the variants in memory at the same time. If not
all of the columns included in the source file are needed, selecting only a
subset of fields can also dramatically decrease the size of the final parquet
file.

```{r}
sql <- glue::glue_sql("
	COPY (
		SELECT *, 
		FROM read_csv({kTab}, AUTO_DETECT = TRUE, NORMALIZE_NAMES = TRUE)
		WHERE Assembly = 'GRCh38' AND _type = 'single nucleotide variant'
	)
	TO {kParquet} (FORMAT 'PARQUET');", .con = con)
DBI::dbExecute(con, sql)
```

## Exploring genetic variants stored in parquet format

I can access the parquet file in different ways from R (or python or other 
environments). For example, I could read its entire content into my R session 
using the
- [{nanoparquet} R package](https://cran.r-project.org/package=nanoparquet), 
or
- [arrow::read_parquet](https://cran.r-project.org/package=arrow) from the
`arrow` package, which supports `dplyr` verbs for downstream processing.

Alternatively, I can use `duckdb`'s internal functionality to connect to
external files, and then pass the database connection to `dbplyr`, which will
translate common `dplyr` verbs into SQL for me. I will demonstrate this
approach in the remainder of this post.

::: {.callout-note collapse="true"}

### duckplyr

Recently, the
[duckplyr R package](https://cran.r-project.org/package=duckplyr)
was released, offering yet another way of using `duckdb` for data analysis.
`duckplyr` was built around an embedded version of `duckdb` to provide a drop-in
replacement for `dplyr's` functionality [^3], including automatic fall-back to
`dplyr` functionality for any functionality that is not supported within
the `duckdb` database instance.

[^3]: With `duckplyr` there is no need to create and manage
an in-memory `duckdb` instance - it's all handled internally.

A main distinction between `duckplyr` and combining `duckdb` with `dbplyr` is
that the former aims to support all R operators, functions, and data types [^4].
It automatically decides when to pull data from the database instance into
the R session (e.g. materialize it in memory). While that is often convenient,
it might require some extract attention, especially when working with very 
large datasets.

[^4]: See [this github issue](https://github.com/tidyverse/duckplyr/issues/145)
for details.

```{r}
suppressPackageStartupMessages(
  library(duckplyr)
)
```

The `duckplyr::read_parquet_duckdb` function connects to a parquet file and
returns a `duckplyr_df` object.

```{r}
#| eval: false 
df <- read_parquet_duckdb(kParquet, prudence = "thrifty")
class(df)
```

```
[1] "prudent_duckplyr_df" "duckplyr_df"         "tbl_df"              "tbl"                
[5] "data.frame"         
```

The function `read_parquet_duckdb()` function accepts a `prudence` argument, 
which governs whether / when data is actually materialized in memory.

- By default, `duckplyr` is `thrifty` and decides at each step of a workflow 
whether the dataset is small enough (< 1 million cells = rows multiplied by 
columns) to materialize in memory right away [^5]. 
- Alternatively, we could choose a `lavish` approach, suitable for small 
datasets, that will trigger the computation be triggered immediately. 
- Finally, setting prudence to `stingy` will avoid automatic materialization 
and requires the use of `collect()` or `as_tibble()` to materialize the data
in a conventional data.frame.

[^5]: Fore more details, please consult `duckplyr's` 
[Memory protection: controlling automatic materialization](https://cran.r-project.org/web/packages/duckplyr/vignettes/prudence.html)
vignette.

:::

### Registering ClinVar as a VIEW

Currently, our in-memory `duckdb` database does not contain any tables.

```{r}
dbListTables(con)
```

Let's select a subset of field from the parquet file and also add a new `width` 
column. Because we will use the output of this query multiple times,
it is useful to add it to the database as a `VIEW`, e.g. a canned query that
is accessible under the `clinvar` name. (Creating a `TABLE` instead of a `VIEW`
would result importing an actual copy the data into the `duckdb` database.)

Whenever we access the `clinvar` view, the SQL query that defines it is 
re-executed, e.g. data is retrieved from the parquet file.

```{sql}
#| connection: con
#| cache: false
#| results: hide
CREATE OR REPLACE VIEW clinvar as
	SELECT cv.*, (cv.end - cv.start) + 1.0 AS width 
	FROM (
		SELECT
			rcvaccession,
			_name as name,
			chromosomeaccession AS seqnames,
			_start AS start,
			stop AS end,
			_type AS type,
			clinicalsignificance,
			clinsigsimple,
			phenotypeids,
			phenotypelist,
			rs_dbsnp AS dbsnp,
			nsvesv_dbvar AS dbvar,
			otherids
		FROM (FROM read_parquet(?kParquet)) AS cv
	) AS cv
```

```{r}
duckdb::dbListTables(con)
```

Next, let's create a toy example of a query, e.g. a genomic region of interest
defined by its genomic coordinates. We could add this query as a 
(potentially temporary) table to our database. Even more conveniently, the 
`duckdb::duckdb_register` function allows us to make the database aware of 
this data.frame without having to transfer any data; it is registered as a
"virtual table" instead.

```{r}
query <- data.frame(
  seqnames = "NC_000016.10",
  start = 2765864,
  end = 2765864
)
query
```

## Querying ClinVar with dplyr

```{r}
suppressPackageStartupMessages(
  library(dbplyr)
)
```

By passing the `duckdb` connection to the `tbl` function, I can use the
database tables in dplyr workflows, e.g. to retrieve any ClinVar variants
with _the same_ coordinates as my query.

```{r}
#| eval: false
dplyr::inner_join(query, tbl(con, "clinvar"),
                  by = join_by(seqnames, start, end))
```

Ouch, that didn't work - it triggered the following error:

```
Error in `auto_copy()`:
! `x` and `y` must share the same src.
ℹ `x` is a <tbl_duckdb_connection/tbl_dbi/tbl_sql/tbl_lazy/tbl> object.
ℹ `y` is a <data.frame> object.
ℹ Set `copy = TRUE` if `y` can be copied to the same source as `x` (may be slow).
```

The `query` data.frame exists in my R session, but the `clinvar` table
is in the database. To join these two sources of information, I need to add
the query to the database first.

I could use the `dbWriteTable` function to create a new (potentially temporary)
table to the database. But there is a more convenient option: the 
`duckdb::duckdb_register` function makes the database aware of 
a data.frame without having to transfer any data; it is registered as a
"virtual table" instead.

```{r}
duckdb::duckdb_register(con, "query", query, overwrite = TRUE)
duckdb::dbListTables(con)
```

Now the join can proceed, and we observe that this SNV corresponds to ClinVar
report `RCV002776799`, e.g. by using `dplyr` to write a JOIN SQL statement
for us.

```{r}
dplyr::inner_join(tbl(con, "query"), tbl(con, "clinvar"),
                  by = join_by(seqnames, start, end))
```

### Querying ClinVar using SQL

We can also write and submit SQL statements directly, and `duckdb` will
search the parquet file (represented by the `clinvar` view) for us.

For example, we can request three rows:

```{sql}
#| connection: con
#| cache: false
SELECT * FROM clinvar LIMIT 3;
```
information about a specific dbSNP identifier (rsid),
```{sql}
#| connection: con
#| cache: false
#| results: hide
SELECT * FROM clinvar WHERE dbsnp == '1294019636';
```
look up a variant by its ClinVar identifier,

```{sql}
#| connection: con
#| cache: false
#| results: hide
SELECT * FROM clinvar WHERE rcvaccession == 'RCV000000014';
```
list all chromosome names that are included in the ClinVar file,

```{sql}
#| connection: con
#| cache: false
#| results: hide
SELECT DISTINCT seqnames AS chr FROM clinvar ORDER BY chr;
```
or retrieve variants include in a specified range (in this case the same
query we made with `dplyr` above).

```{sql}
#| connection: con
#| cache: false
#| results: hide
SELECT * FROM clinvar
INNER JOIN query
ON (
	clinvar.seqnames = query.seqnames AND
	clinvar.start <= query.end AND
	clinvar.end >= query.start
	)
```

Finally, we close the connectsion to the in-memory database.

```{r}
duckdb::dbDisconnect(con)
```

## Conclusions

Storing large datasets in parquet files and querying them using `duckdb` is one
of my favorite things I learned this year. In this example, I am using a local 
in-memory duckdb instance to search a local file, but I could do the same with
files stored remotely e.g. on AWS S3. 

Eventually, as file sizes increase, it will become too slow to query files over
the network, and I will have to consider moving my query engine closer to the
data, e.g. by running `duckdb` on a cloud instance, or e.g. by querying the
parquet files through a cloud database such as 
[motherduck](https://motherduck.com) or
[AWS Athena](https://tomsing1.github.io/blog/posts/athena/index.html#creating-an-external-table-with-aws-athena). But or prototyping and medium sized datasets, `duckdb` is a 
powerful addition to my genetics toolkit.

## Reproducibility

<details>
<summary>
Session Information
</summary>

```{r}
sessionInfo()
```

</details>
