---
title: "anndataR: converting single-cell data, from R to python and back"
author: "Thomas Sandmann"
date: "2026-02-08"
execute:
  echo: true
  warning: false
  message: false
  cache: false
freeze: true
categories: [R, TIL]
editor:
  markdown:
    wrap: 72
format:
  html:
    anchor-sections: true
    toc: true
    toc-depth: 4
    code-tools:
      source: true
      toggle: false
      caption: none
editor_options: 
  chunk_output_type: console
---

## tl;dr

Both the R and python communities have created amazing software tools to analyze
single-cell data. It used to be tricky to convert between the most prominent
serialized data structures: `SingleCellExperiment` and `Seurat` R objects and
`AnnData` python objects. Luckily, Bioconductor release 3.22 introduced the
[anndataR R package](https://bioconductor.org/packages//release/bioc/html/anndataR.html),
authored by 
[Robrecht Cannoodt](https://cannoodt.dev),
[Luke Zappia](https://lazappi.id.au),
[Martin Morgan](https://mtmorgan.github.io) and
[Louise Deconinck](https://orcid.org/0000-0001-8100-6823), 
which offers a reliable solution [^1].

[^1]: For a comparison with other R packages that offer similar functionality,
check out the vignettes that accompany the `anndataR` package, for example
[this one](https://bioconductor.org/packages//release/bioc/vignettes/anndataR/inst/doc/anndataR.html).

## Lin et al's single-cell RNA-seq datasets

In 2021, 
[Lin et al](https://pubmed.ncbi.nlm.nih.gov/34358451/) 
published three separate single-cell RNA-seq datasets 
that characterize the transcriptomes of neurons differentiated from induced
pluripotent stem cells (iPSC) through overexpression of the NGN2 transcription
factor. The authors made the data available as serialized Seurat objects 
(version 3.1), available for download from 
[Mendeley Data](https://data.mendeley.com/datasets/y3s4hnyvg6/1).

```{r}
suppressPackageStartupMessages({
  library(anndataR)
  library(fs)
  library(rhdf5)
  library(Seurat)
})
options(timeout=3600)  # avoid timeouts while downloading large files
```

Let's start with a function that allows us to download a deposited file and
store it in a temporary location on the local system.

<details>
<summary>
Download function
</summary>

```{r}
#' Download a serialized Seurat object published by Lin et al from Mendeley data
#'
#' @param experiment Scalar character, the experiment to download
#' @return Scalar character, the path to the downloaded file
#' @details The Seurat files were referenced in the
#' [preprint](https://www.biorxiv.org/content/10.1101/2020.11.19.389445v1.full)
#' and it is possible that the final results published in Stem Cell Reports
#' are (slightly) different. Available serialized Seurat objects:
#' - `d14`:
#'   - scRNA-seq data of Ngn2-iNeuron at d14, varied Dox duration during culture
#'   - RDS file size: 284.3 MB
#'   - 51723 features across 2767 cells:
#     - iN_Dox_ctrl: 1351 cells
#     - iN_Dox_d1: 311 cells
#     - iN_Dox_d3: 378 cells:
#     - iN_Dox_d5: 727 cells
#' - `w5`:
#'   - scRNA-seq data of w5 Ngn2-iNeuron from different cell types and clones
#'   - RDS file size: 114 MB
#'   - 17546 features across 3866 cells:
#     - line 09b2: 993 cells
#     - line 409b2_clone: 1654 cells
#     - line sc102a1: 1219 cells
#' - `timecourse`:
#'   - Time-course scRNA-seq data of Ngn2-induced neuron differentiation
#'   - RDS file size: 1454.1 MB
#'   - 17546 features across 29554 cells:
#     - h0: 1636 cells
#     - h6/12: 6874 cells
#     - d1: 3148 cells
#     - d2: 6350 cells
#     - d5: 1907 cells
#     - w2: 1508 cells
#     - w4: 7141 cells
#     - w5: 990 cells
download_seurat_object <- function(experiment = c("d14", "w5", "timecourse")) {
  experiment <- match.arg(experiment)
  root <- "https://data.mendeley.com/public-files/datasets/y3s4hnyvg6/files/"
  url <- switch(
    experiment,
    "d14" = paste0(
      root,
      "b33734ff-8f10-4bf1-8bde-d7af79f4177f/file_downloaded"
    ),
    "w5" = paste0(
      root,
      "c3b21223-bd13-4bd4-82e3-1cb143136244/file_downloaded"
    ),
    "timecourse" = paste0(
      root,
      "27ec7771-eda0-44d4-9ae5-0859e26fabb5/file_downloaded"
    )
  )
  temp_file <- tempfile(fileext = ".rds")
  download.file(url, destfile = temp_file)
  return(temp_file)
}
```

</details>

For example, we can retrieve the `w5` dataset, containing data from 
Ngn2-iNeurons derived from different cell types and clones, and then load it 
into our R session. Because the definition of Seurat objects has been updated 
since the authored deposited the files, we update it on the fly with 
`Seurat::UpdateSeuratObject`.

::: {.callout-wqrning}

The following call will download 114.3 MB of data.

:::

```{r}
#| warning: false
seurat_obj <- download_seurat_object('w5') |>
  readRDS() |>
  Seurat::UpdateSeuratObject()
```

Next, let's take a look at the Seurat object itself. It contains data from two 
`assays`: "RNA" and "integrated". Because we can only store results from a 
single assay in an `annData` object, we will focus on the `RNA` assay here.

```{r}
names(seurat_obj@assays)
```

The Seurat object includes multiple metadata columns that provide information
about each cell in the dataset.

```{r}
colnames(seurat_obj@meta.data)
```

In addition, the cell identity vector assigns each cell to one of 
`r nlevels(Idents(seurat_obj))` clusters. Let's add this information to the
`metadata` slot [^2].

[^2]: The `integrated_snn_res.0.3_merged` metadata column contains _almost_ 
the same information as the cell identities, but it only defines `r length(unique(seurat_obj$integrated_snn_res.0.3_merged))` clusters: 
<pre>
`r paste(capture.output(table(seurat_obj$integrated_snn_res.0.3_merged, Idents(seurat_obj))), collapse = "\n")`
</pre>

The object also contains the reduced dimensions the authors calculated in the
`pca` and `umap`. Let's plot the cells along their UMAP coordinates and color
them by their assigned clusters:

```{r}
#| fig-width: 6
#| fig-height: 5
seurat_obj$seurat_clusters <- Seurat::Idents(seurat_obj) 
Seurat::DimPlot(seurat_obj)
```

## Converating from Seurat to an AnnData object

Converting the Seurat object to an `InMemoryAnnData` R object is as simple
as calling `anndataR::as_AnnData`. By default, the function will try to guess
which slots of the Seurat object should be mapped to the corresponding slots 
of the `annData` object. Here, we specify the layers, reduced dimensions 
(`pca` and `umap`) and nearest neighbor graphs explicitely, to make sure they
are correctly included in the converted object.

```{r}
adata <- anndataR::as_AnnData(
  x = seurat_obj,
  assay_name = "RNA",
  layers_mapping = c(
    counts = "counts",
    data = "data"
  ),
  x_mapping = "counts",
  obsp_mapping = c(
    integrated_nn = "integrated_nn",
    integrated_snn = "integrated_snn"
  ),
  obsm_mapping = c(X_pca = "pca", X_umap = "umap"),
  output_class = "InMemory"
)
adata
```

The `adata` object mirrors the structure of the python [AnnData](https://anndata.readthedocs.io/en/stable/generated/anndata.AnnData.html)
class, defined in the 
[anndata python module](https://anndata.readthedocs.io/en/stable/index.html).

By default, the data is kept in memory, e.g. the `output_class` is set to 
`InMemory`. Alternatively, we can request `HDF5AnnData` or 
`ReticulateAnnData` output objects.

::: {.callout-info}

- A `ReticulateAnnData` object wraps a python `anndata.AnnData` object using 
[reticulate](https://rstudio.github.io/reticulate/). 
This allows direct interaction with python `AnnData` objects while maintaining
the R interface, but it requires the configuration of a python virtual
environment.
- An `HDF5AnnData` class stores the assay data inside an HDF5 file on disk; 
to request this output class, a `file` argument is specified. This class 
provides an interface to a H5AD file and minimal data is stored in memory 
until it is requested. This is a great option when the Seurat object is very
large, and we don't want to load the entire dataset into memory at once. 

:::

Because this dataset is relatively small, we can load it into memory first
(see above) and then write it to disk as an `h5ad` file:

```{r}
#| warning: false
#| message: false
h5ad_file_path <- file.path(tempdir(), "anndata.h5ad")
anndataR::write_h5ad(
  object = adata,
  compression = "gzip",
  path = h5ad_file_path
)
fs::file_info(h5ad_file_path)[, c("path", "type", "size")]
```

Alternatively, e.g. for large datasets, we can specify the 
`output_class = "HDF5AnnData", file = hd5a_file_path` arguments. To demonstrate,
we remove the h5ad file we create above, and write it to disk directly. The
returned object is an `HDF5AnnData`:

```{r}
fs::file_delete(h5ad_file_path)
adata <- anndataR::as_AnnData(
  x = seurat_obj,
  assay_name = "RNA",
  layers_mapping = c(
    counts = "counts",
    data = "data"
  ),
  x_mapping = "counts",
  obsp_mapping = c(
    integrated_nn = "integrated_nn",
    integrated_snn = "integrated_snn"
  ),
  obsm_mapping = c(X_pca = "pca", X_umap = "umap"),
  output_class = "HDF5AnnData", 
  file = h5ad_file_path
)
adata
```

The `anndataR` package not only simplifies the conversion between different
single-cell objects, but it also enables R users to interact with `h5ad` files
directly - awesome!

## Reproducibility

<details>
<summary>
Session Information
</summary>

```{r}
sessionInfo()
```

</details>
