---
title: "Figure size, layout & tabsets with Quarto"
author: "Thomas Sandmann"
date: "2022-12-22"
freeze: true
categories: [TIL, R, quarto]
editor: 
  markdown: 
    wrap: 72
---

In this document, I am experimenting with various attributes
that organize the layout, size and placement of figures of
Quarto document. For more details, please check out the
[official documentation](https://quarto.org/docs/guide/),
especially the topics on 
[figures](https://quarto.org/docs/authoring/figures.html)
and
[article layout](https://quarto.org/docs/authoring/article-layout.html).

::: {.callout-note}

For illustration, I am displaying both the code that generates 
a simple plot as well as the attributes that determine how
it is rendered, e.g. the `:::` tags interspersed with the
code blocks, and the `#|` attributes within individual
code cells. See the documentation on 
[executable blocks](https://quarto.org/docs/output-formats/html-code.html#executable-blocks)
for details.
:::

First, let's generate a simple plot, so we can see the effect of different
attributes on how it is rendered in subsequent code cells.

To start, we render the output without specifying any custom attributes, e.g.
using the default settings for this Quarto website:

```{r}
library(ggplot2)
theme_set(theme_linedraw(base_size = 14))
p <- ggplot(mtcars, aes(x = mpg, y = drat)) + 
  geom_point(color = "skyblue", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", formula = 'y ~ x', se = FALSE) +
  theme(panel.grid = element_blank())
p
```

### Width and height of individual figures

The `fig-width` and `fig-height` attributes specify the 
dimensions of the image file that is generated. The `out-width` attribute determines the size at which that
image is displayed in the rendered HTML page.

```
#| fig-width: 4
#| fig-height: 5
#| out-width: "50%"
#| fig-align: "center"
```

```{r}
#| fig-width: 4
#| fig-height: 5
#| out-width: "50%"
#| fig-align: "center"
p
```

For example, the same image can be displayed at 50% of the
width of the enclosing `<div>`.

```
#| fig-width: 4
#| fig-height: 5
#| out-width: "25%"
#| fig-align: "center"
```

```{r}
#| fig-width: 4
#| fig-height: 5
#| out-width: "25%"
#| fig-align: "center"
p
```

### Layout: columns and rows

The `layout-ncol` and `layout-nrow` attributes govern the placement 
of multiple figures within the same element. For example, we
can place two figures next to each other, in two column. 

The `fig-align` attributes specify the figure alignment within each column. 

::: {.callout-tip}
The `out-width` attribute is always relative to its
enclosing element, e.g. here `out-width: "50%"` refers to
half of the width of a *column*, not the page width. 
:::

```
::: {layout-ncol=2}
```

::: {layout-ncol=2}

```
#| out-width: "50%"
#| fig-align: "center"
```

```
#| out-width: "30%"
#| fig-align: "right"
```

```{r}
#| out-width: "50%"
#| fig-align: "center"
p
```


```{r}
#| out-width: "30%"
#| fig-align: "right"
p
```

:::

```
:::
```

### Tabsets

[Tabsets](https://bookdown.org/yihui/rmarkdown-cookbook/html-tabs.html) 
can be used to organize contents, e.g. by hiding content
until the other clicks on the tab's header.

The layout of the first tabset contains just one column and 
row. 

```
::: {.panel-tabset}
```

::: {.panel-tabset}

### panel 1

```{r}
#| out-width: "50%"
#| fig-align: "center"
p
```

### panel 2

The second panel is subdivided into two columns. (Note the 
use of the `::::` tag, nested within the `:::` parent tag.)

```
:::: {layout-ncol=2}
```
:::: {layout-ncol=2}

```{r}
p
```

```{r}
p
```

```
::::
```

::::
:::

```
:::
```