---
title: "Full text search in Postgres - the R way"
author: "Thomas Sandmann"
date: "2022-12-12"
freeze: true
categories: [TIL, R, postgres]
editor: 
  markdown: 
    wrap: 72
---

I have been learning how to organize, search and modify data in a 
[Postgres](https://www.postgresql.org/) 
database by working through 
[Anthony DeBarros'](https://www.wsj.com/news/author/anthony-debarros) 
excellent book 
[Practical SQL](https://nostarch.com/practical-sql-2nd-edition).

Because I currently perform most of my data analyses in R, I am using the 
great 
[RPostgres](https://cran.r-project.org/package=RPostgres),
[DBI](https://cran.r-project.org/package=DBI)
and
[glue](https://cran.r-project.org/package=glue)
packages to interface with Postgres - without ever leaving my R session. 

Today I learned how to create a full text search index and how to search it
with one or more search terms.

## Connecting to Postgres

For this example, I created a toy database `full_text_search` in my local
Postgres server. I connect to it with the `DBI::dbConnect` command, and by 
passing it the `RPostgres::Postgres()` driver.

```{r}
library(DBI)
library(glue)
library(RPostgres)
library(sessioninfo)

# Connect to a (prexisting) postgres database called `full_text_search`
con <- DBI::dbConnect(
  dbname = "full_text_search",
  drv = RPostgres::Postgres(),
  host = "localhost",
  port = 5432L,
  user = "postgres"
  )
```

Because this is a toy example, I start with a fresh table `datasets`. (In case
it already exists from previous experimentation, I drop the table if necessary).

Let's define four fields for the table:

- `id`: the unique identifier
- `name`: the short name of each entry
- `title`: a longer title
- `description`: a paragraph describing the entry
- `created`: a date and time the entry was added to the database 

```{r}
# drop the `datasets` table if it already exists
if (DBI::dbExistsTable(con, "datasets")) DBI::dbRemoveTable(con, "datasets")

# create the empty `datasets` table
sql <- glue_sql("
      CREATE TABLE IF NOT EXISTS datasets (
      id bigserial PRIMARY KEY,
      name text,
      title text,
      description text,
      created timestamp with time zone default current_timestamp not null
    );", .con = con)
res <- suppressMessages(DBI::dbSendStatement(con, sql))
DBI::dbClearResult(res)
DBI::dbReadTable(con, "datasets")
```

Initially, our new database is empty. Let's populate them with three entries,
each describing a popular dataset shipped with R's built-in
[datasets](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html)
package.

```{r}
# some example entries
buildin_datasets <- list(
  mtcars = list(
    "name" = "mtcars", 
    "title" = "The built-in mtcars dataset from the datasets R package.",
    "description" = gsub(
      "\r?\n|\r", " ", 
      "The data was extracted from the 1974 Motor Trend US magazine, and 
comprises fuel consumption and 10 aspects of automobile design and
performance for 32 automobiles (1973–74 models).")
  ), 
  airmiles = list(
    name = "airmiles",
    title = "The built-in airmiles dataset from the datasets R package",
    description = gsub(
      "\r?\n|\r", " ", 
      "The revenue passenger miles flown by commercial airlines in the United
States for each year from 1937 to 1960.")
  ),
  attitude = list(
    name = "attitude", 
    title = "The built-in attitude dataset from the datasets R package",
    description = gsub(
      "\r?\n|\r", " ", 
      "From a survey of the clerical employees of a large financial
organization, the data are aggregated from the questionnaires of the
approximately 35 employees for each of 30 (randomly selected) departments. 
The numbers give the percent proportion of favourable responses to seven
questions in each department.")
  )
)
```

Next, we loop over each element of the list and use the `glue_sql()` command
to unpack both the names (`names(dataset)`) and the values of each field for 
this entry. Then we update the `datasets` table with this new information.

Afterward, we retrieve the `name` and `title` fields to verify the
correct import:

```{r}
for (dataset in buildin_datasets) {
  sql <- glue_sql(
    "INSERT INTO datasets ({`names(dataset)`*})
   VALUES ({dataset*});", 
    .con = con)
  res <- suppressMessages(DBI::dbSendStatement(con, sql))
  DBI::dbClearResult(res)
}
DBI::dbGetQuery(con, "SELECT name, title from datasets;")
```

My goal is to enable full-text search for the `description` field. First, we 
need to add a `tsvector` field and populate it with the tokenized contents 
of each `description`.

```{r}
# create a column to hold tokens for full text search
sql <- glue_sql(
  "ALTER TABLE datasets
   ADD COLUMN search_description_text tsvector;", 
  .con = con)
res <- suppressMessages(DBI::dbSendStatement(con, sql))
DBI::dbClearResult(res)
DBI::dbListFields(con, "datasets")
```

At this point, the `search_description_text` field is still empty. Let's 
copy the `descriptions` into it - and tokenize them at the same time. For 
illustration, we retrieve the tokens for the first dataset:

```{r}
# copy the description into search tokens
sql <- glue_sql(
  "UPDATE datasets
   SET search_description_text = to_tsvector('english', description);", 
  .con = con)
DBI::dbExecute(con, sql)
DBI::dbGetQuery(con, 
                "SELECT name, search_description_text from datasets LIMIT 1;")
```

To speed up the full-text search, we add a
[Generalized Inverted Index (GIN)](https://www.postgresql.org/docs/current/gin-intro.html#:~:text=GIN%20stands%20for%20Generalized%20Inverted,appear%20within%20the%20composite%20items.)
index for the `search_description_text` column as well:

```{r}
# create the search index
sql <- glue_sql(
  "CREATE INDEX search_description_idx
   ON datasets
   USING gin(search_description_text);",
  .con = con)
DBI::dbExecute(con, sql)
```

Now we are ready to perform our first search. Let's look up the term `data` in
the `description` fields. (The `ts_headline` command returns the location of the
match, e.g. the context it was found in.)

```{r}
# search the description field and show the matching location
term <- "data"
sql <- glue_sql(
  "SELECT id, name,
  ts_headline(description, to_tsquery('english', {term}),
     'StartSel = <,
      StopSel = >,
      MinWords = 5,
      MaxWords = 7,
      MaxFragments = 1')
  FROM datasets
  WHERE search_description_text @@ to_tsquery('english', {term})
  ORDER BY created;",
  .con = con)
res <- suppressMessages(DBI::dbSendStatement(con, sql))
DBI::dbFetch(res)
DBI::dbClearResult(res)
```

We can also combine search terms, e.g. searching for either `employee` _or_ 
`motor` terms:

```{r}
# using multiple search terms
term <- "employee | motor"  # OR
sql <- glue_sql(
  "SELECT id, name,
  ts_headline(description, to_tsquery('english', {term}),
     'StartSel = <,
      StopSel = >,
      MinWords = 5,
      MaxWords = 7,
      MaxFragments = 1')
  FROM datasets
  WHERE search_description_text @@ to_tsquery('english', {term})
  ORDER BY created;",
  .con = con)
res <- suppressMessages(DBI::dbSendStatement(con, sql))
DBI::dbFetch(res)
DBI::dbClearResult(res)
```

Similarly, we can narrow our search by requiring both `data` _and_ `employee`
terms to appear in the same description:

```{r}
term <- "data & employee"  # AND
sql <- glue_sql(
  "SELECT id, name,
  ts_headline(description, to_tsquery('english', {term}),
     'StartSel = <,
      StopSel = >,
      MinWords = 5,
      MaxWords = 7,
      MaxFragments = 1')
  FROM datasets
  WHERE search_description_text @@ to_tsquery('english', {term})
  ORDER BY created;",
  .con = con)
res <- suppressMessages(DBI::dbSendStatement(con, sql))
DBI::dbFetch(res)
DBI::dbClearResult(res)
```

That's it. Thanks again to 
[Anthony DeBarros'](https://www.wsj.com/news/author/anthony-debarros) 
for his excellent introduction to 
[Practical SQL](https://nostarch.com/practical-sql-2nd-edition)!

<details>
```{r}
sessioninfo::session_info()
```
</details>