---
title: "Dipping my toes into the ducklake: Exploring gene expression data with R and python"
author: "Thomas Sandmann"
date: "2025-07-10"
execute:
  echo: true
  warning: false
  message: false
  cache: true
freeze: true
categories: [R, duckdb, python, TIL]
editor:
  markdown:
    wrap: 72
format:
  html:
    anchor-sections: true
    toc: true
    toc-depth: 4
    code-tools:
      source: true
      toggle: false
      caption: none
editor_options: 
  chunk_output_type: inline
---

## tl;dr

This week I learned how to store 1) the gene-level results (raw and
normalized counts) and 2) the results of multiple comparisons between sample
groups (p-values, log2 fold change, etc) in a
[data lake](https://en.wikipedia.org/wiki/Data_lake), specifically a `ducklake`
managed by `duckdb`. It allows me to store large tables in parquet files on
disk (or in the cloud), and to interact with them from R, python or via plain 
SQL.

::: {.callout-note}
### Data lakes 💧
A data lake describes a setup where (usually large) datasets are not added as
tables to the relational database, but are stored in files instead. These files
can be stored on the local system or in the cloud, and are queried as needed.
Different file formats are used, including e.g. CSV or
[parquet](https://parquet.apache.org/docs/file-format/).

Keeping track of external files often requires additional infrastructure, e.g. a
[catalog](https://www.ibm.com/think/topics/data-catalog), and updating /
changing data while keeping track of file locations and names can pose
challenges. The 
[DuckLake release post](https://duckdb.org/2025/05/27/ducklake.html) 
provides a great summary and explains the motivation behind the development of
the extension.

:::

## Managing gene expression data across studies

Over the years, I have been involved in planning, executing and analyzing 
hundreds of RNA-seq experiments. First, I focus on a single study, performing
exploratory and differential expression analyses to answer key scientific
questions. Combining measurements (e.g. gene- or exon-level counts), sample and
feature metadata in a single object, e.g. a 
[SummarizedExperiment R object](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html)
or an 
[AnnData python object](https://anndata.readthedocs.io/en/stable/)
greatly facilitates these study-level analyses.

But sooner or later, additional questions arise that require the analysis
of _multiple_ experiments, e.g. comparing the results of different studies 
performed using the same model system, employed the same perturbation or 
were conducted at different time points. Now, I need to wrangle multiple
_different datasets_ - each of which might contain multiple different
experimental groups that can be compared to each other.

For the longest time, I have stored data from individual studies as serialized
R objects, e.g. `DGEList` or `MArrayLM` objects with RNA-seq counts and
results of statistical comparisons, respectively. For cross-study comparisons I
would load each of the required objects into an analysis session, extract the
statistics of interest, stack them into a table, etc - writing lots of bespoke
code for each analysis.

In many cases, it is of interest to compare the _results_ of differential
expression analysis, e.g. the observed fold changes and associated statistics 
(p-values, FDR, confidence intervals, etc). What if I extracted the relevant 
statistics as soon I first analysed a new study, and then stored them in
a standardized, language-agnostic format? Over time, this collection of results
would grow and could facilitate even complex cross-study analyses were 
needed[^1].

[^1]: Often, studies are carefully designed to support specific comparisons,
e.g. they contain treatment and matched control groups who were assessed on
the same day or within the same processing batches. This information is crucial
for the analysis of treatment effects, and well-designed experiments can
remove the unwanted variation introduced by batch effects. That's why it is
often more useful to contrast the differential expression results (e.g. the 
_changes_ observed between treatment and control conditions) recorded in two
independent experiments than the trying to compare the raw measurements
themselves (as each of the studies carries its own complement of batch effects,
technical artifacts, etc).

This flowchart illustrates my workflow for a single study, but it could
be repeated for each new RNA-seq experiment I encounter:

```{mermaid}
%%| echo: false
flowchart LR
  Samples["Samples<br/>━━━━━<br/>A1,A2,A3<br>B1,B2,B3<br>C1,C2,C3"] --> Counts(RNAseq)
  Counts --> NormCounts("Normalized counts")
  NormCounts --> Results(Counts, CPMs)
  NormCounts --> Comparisons("Comparisons<br/>━━━━━<br/>A vs B<br>B vs C<br>A vs C")
  Comparisons --> Stats(Fold change, p-values)
  Results --> ducklake[(ducklake 🦆)]
  Stats --> ducklake
```

```{r}
#| message: false
#| warning: false
#| cache: false
suppressPackageStartupMessages({
  library(DESeq2)
  library(dbplyr)
  library(dplyr)
  library(duckdb)
  library(DT)
  library(edgeR)
  library(fs)
  library(ggplot2)
  library(glue)
  library(limma)
  library(sessioninfo)
  library(stringr)
  library(tidyr)
})
```

## Differential gene expression data: Wang et al, 2018

Let's get some gene expression data to play with. In 2018, 
[Wang et al](https://www.ncbi.nlm.nih.gov/pubmed/30558641) 
published the transcriptome analysis of mouse microglia, isolated either from 
the 
[rTg4510 mouse model](https://www.alzforum.org/research-models/rtgtaup301l4510) 
or wildtype animals. Four age groups of mice (2-, 4-, 6-, and 8-months) were 
analyzed to capture longitudinal gene expression changes that correspond to 
varying levels of pathology, from minimal tau accumulation to massive neuronal
loss.

A `DESeqDataset` object with raw counts, sample and gene annotations is
available in my [rnaseq-examples](https://github.com/tomsing1/rnaseq-examples) 
github repository[^2].

[^2]: The repository contains an installable R package, but here I am just 
retrieving the serialized R object and load it into the R session.

::: {.callout-note collapse="true"}
#### Retrieving the raw RNA-seq counts 🧬

```{r}
#| message: false
#| warning: false
dge <- local({
  temp_file <- tempfile(fileext = ".rda")
  download.file(url = paste0("https://github.com/tomsing1/rnaseq-examples/raw/",
                             "refs/heads/main/data/tau.rda"), 
                destfile = temp_file
  )
  load(temp_file)
  edgeR::normLibSizes(tau, method = "TMM")
})
```
:::

There are 4 animals in each of the `r nlevels(dge$samples$group)` experimental 
groups.

```{r}
table(dge$samples$group)
```

We can compare the different genotypes (and time points) e.g. using the [limma/voom](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29) workflow. Because there four different time points (2, 4, 6 and 8 months), we 
extract four different contrasts, each comparing Tau-expressing with wildtype 
animals of the same age.

::: {.callout-note collapse="true"}
#### Differential expression analysis with limma / voom 📊

```{r fig.width=5, fig.height=5}
design <- model.matrix(~ 0 + group, data = dge$samples)
colnames(design) <- sub("group", "", colnames(design))

keep <- filterByExpr(dge, design = design)
fit <- voomLmFit(
  counts = dge[keep & dge$genes$gene_type == "protein_coding", ],
  design = design,
  plot = TRUE,
  sample.weights = TRUE)

contrasts = makeContrasts(
  two_months = "rTg4510_2 - WT_2",
  four_months = "rTg4510_4 - WT_4",
  six_months = "rTg4510_6 - WT_6",
  eight_months = "rTg4510_8 - WT_8",
  levels = design
)
fit2 <- contrasts.fit(fit, contrasts = contrasts)
fit2 <- eBayes(fit2)
```
:::

At each time point, we detect a large number of differentially expressed genes
(with the default FDR \< 5% threshold).

```{r}
summary(decideTests(fit2))
```

::: {.callout-note collapse="true"}
#### Statistical results 📈

The `fit2` MArrayLM object contains statistics for all of the contrasts, and 
this information can be extracted with the `limma::topTable` function.

```{r}
#| cache: false
stats <- lapply(colnames(fit2), \(contrast) {
  tt <- topTable(fit2, coef = contrast, confint = TRUE, number = Inf)[
    , c("symbol", "gene_id", "logFC", "CI.L", "CI.R", "P.Value", "adj.P.Val")]
  data.frame(
    contrast = contrast, 
    tt
  )
})
stats <- do.call(rbind, stats)   |>
  dplyr::mutate(study = "Wang_2018", .before = 1)
row.names(stats) <- NULL
head(stats)
```

We might also be interested in the raw counts and normalized expression scores
(Counts per million reads) for each gene in each of the samples.

```{r}
#| cache: false
exprs <- local({
  counts <- dge$counts |>
    as.data.frame() |>
    dplyr::bind_cols(dge$genes) |>
    tidyr::pivot_longer(cols = starts_with("DRN-"), names_to = "sample_id", 
                        values_to = "counts") |>
    dplyr::pull("counts")
  
  edgeR::cpm(dge, normalized.lib.sizes = TRUE) |>
    as.data.frame() |>
    dplyr::bind_cols(dge$genes) |>
    tidyr::pivot_longer(cols = starts_with("DRN-"), names_to = "sample_id", 
                        values_to = "cpm") |>
    dplyr::mutate(counts = counts) |>
    dplyr::left_join(
      data.frame(sample_id = colnames(dge), 
                 group = as.character(dge$samples$group)
      ),
      by = "sample_id"
    )
}) |>
  dplyr::mutate(study = "Wang_2018", .before = 1)
head(exprs)
```
:::

We now have two large tables:

- `exprs`: `r nrow(exprs)` rows and `r ncol(exprs)` columns with raw and 
normalized counts for each of the `r ncol(dge)` samples.
- `stats`: `r nrow(stats)` rows and `r ncol(stats)` columns with the results of
the differential expression analysis, across `r ncol(fit2)` contrasts.

To make these tables easier to access from other tools (not just within R), 
let's create our first `ducklake`!

## Instantiating a DuckLake 🦆

In May 2025, the `duckdb` team released the 
[DuckLake core extension](https://duckdb.org/docs/stable/core_extensions/ducklake) 
to manage a data lake.

A `DuckLake` takes advantage of a *catalog database*, and we can choose among 
many different options 
[as outlined here](https://ducklake.select/docs/stable/duckdb/usage/choosing_a_catalog_database). 
For this example, I will use `duckdb` as the database engine, because there will
only be a single client (e.g. my R session) that will interact with it.

All interactions with my `DuckLake` will be conducted using SQL, e.g. it is
not tied to R (or python, etc). Here, I am leveraging the 
[duckdb R package](https://duckdb.r-universe.dev/duckdb)
to connect to duckdb from within R, but the same functionality would be 
available e.g. by interacting with `duckdb` via the command line.

We need version 1.3.1 of `duckdb` or above:

```{r}
package.version("duckdb")  #  <1> 
```

First, we define two paths:

```{r}
#| cache: false
db_file <- file.path(tempdir(), "metadata.ducklake") # <1>
work_dir <- file.path(tempdir(), "lake_data_files")  # <2>
```

1. The location of the metadata database. Here, we are using a `duckdb` 
  instance, but we could also use another relational database (e.g. Postgres).
2. The path to a folder that will contain the data in the form of parquet files.

```{r}
#| include: false
#| cache: false
if (file.exists(db_file)) unlink(db_file)
if (dir.exists(work_dir)) unlink(work_dir, recursive = TRUE)
```

```{r}
#| cache: false
#| results: hide
con <- dbConnect(duckdb())  # <1>
db_path <- glue_sql('{paste0("ducklake:", db_file)}', .con = con)  # <2>
```

1.  We instantiate a `duckdb` instance in memory and connect to it, to manage 
  the connection to the metadata database (which also happens to be a `duckdb`
  instance, but a different one!)
2.  The path to the data lake's metadata database is prefixed with`ducklake:`.

The following SQL code first installs the `ducklake` extension[^3], which is
then used automatically when a `DuckLake` is attached by

1.  Specifying the metadata dataase (by pointing to the `db_path` location) and
2.  The location where data is stored in parquet files.

[^3]: See the 
[ducklake documentation](https://ducklake.select/docs/stable/duckdb/introduction)
for details!

Afterwards, we can work with the `my_ducklake` data lake in the same way we 
interact with any other database, e.g. we can create tables (`exprs` and 
`stats`) and copy the gene expression measurements and differential expression
statistics into them.

```{sql}
#| connection: con
#| cache: false
#| results: hide
INSTALL ducklake;
ATTACH ?db_path AS my_ducklake (DATA_PATH ?work_dir);
USE my_ducklake;
```

Let's start by copying the results of two of the comparisons into the `stats`
table:

```{r}
#| cache: false
#| warning: false
dbWriteTable(  # <1>
  conn = con, 
  name = "stats",
  value = dplyr::filter(stats, contrast %in% c("two_months", "four_months"))
)
dbWriteTable(con, "exprs", exprs) # <2>
```

1. The `dbWriteTable` function creates a table and inserts the content
  of a data.frame. First, we add only the results for two contrasts (by 
  filtering the `stats` data.frame.)
2. The second call to `dbWriteTable` inserts the full `exprs` table with
  the raw and normalized expression scores for each gene in each sample.

We can confirm that statistics for ~ 14,000 genes are now available:

```{sql}
#| connection: con
#| cache: false
SELECT contrast, COUNT(symbol) FROM my_ducklake.stats GROUP BY contrast;
```

For example, we can retrieve the differential expression results for a 
single gene, _Cd34_:

```{r}
#| cache: false
tbl(con, "stats") |>
  dplyr::filter(symbol == "Cd34")
```

Behind the scenes, `duckdb` has stored our data in two parquet files, one for
each table:

```{sql}
#| connection: con
#| cache: false
FROM ducklake_table_info('my_ducklake')
```

The files themselves are stored in the `work_dir` directory we specified when we
instantiated the `ducklake`. (Here, I am using a folder on the local system, but
I could have provided a cloud location on AWS S3 or within a Google Cloud bucket
instead.) The parquet file for the  `stats` table is in the eponymous 
sub-directory:

```{r}
#| cache: false
fs::dir_info(file.path(work_dir, "main", "stats")) |>
  dplyr::mutate(filename = basename(path)) |>
  dplyr::select(filename, size)
```

Next, we append the differential expression statistics for the remaining two
comparisons to the same `stats` table in our `ducklake`:

```{r}
#| cache: false
#| warning: false
dbAppendTable(
  conn = con, 
  name = "stats",
  value = dplyr::filter(stats, contrast %in% c("six_months", "eight_months"))
)
```

Inspecting the `stats` directory now reveals that a second parquet has been
created:

```{r}
#| cache: false
fs::dir_info(file.path(work_dir, "main", "stats")) |>
  dplyr::mutate(filename = basename(path)) |>
  dplyr::select(filename, size)
```

and we now query results for all four comparisons:

```{sql}
#| connection: con
#| cache: false
SELECT contrast, COUNT(symbol) FROM my_ducklake.stats GROUP BY contrast;
```

::: {.callout-note collapse="true"}
## Time travel 🧭

Data can change over time. For example, what if we decided to include an 
additional covariate in our linear models? We would obtain an updated set of
differential expression statistics, and could replace the old results in our
`ducklake`. But what if a previous analysis used the old results? How could 
we track the changes that were made over time - or even choose whether we want
to use the old or the new results for a report?

Actually, `ducklake` tracks all modifications that are made, including 
schema changes or insertions / deletions or update of data. Every modification
triggers the creation of a new 
[snapshot](https://ducklake.select/docs/stable/duckdb/usage/snapshots):

```{sql}
#| connection: con
#| cache: false
SELECT * FROM ducklake_snapshots('my_ducklake');
```

We can use the `AT` keyword to query the `ducklake` at any point in time.
For example, we can query the latest snapshot of the database, e.g. the third
snapshot:

```{sql}
#| connection: con
#| cache: false
SELECT contrast, COUNT(symbol) 
FROM my_ducklake.stats AT (VERSION => 3)
GROUP BY contrast;
```

But we can also issue the same query to inspect the number of data points
that were available after the second snapshot was taken:

```{sql}
#| connection: con
#| cache: false
SELECT contrast, COUNT(symbol) 
FROM my_ducklake.stats AT (VERSION => 2)
GROUP BY contrast;
```

Alternatively, we can also provide a timestamp or the time interval of interest,
please check the
[ducklate documentation](https://ducklake.select/docs/stable/duckdb/usage/time_travel.html)
for more examples.

The `ducklake` extension also provides helper functions, e.g. a call to the 
`ducklake_table_changes` function can highlight changes that occured
between snapshots 1 and 2:

```{sql}
#| connection: con
#| cache: false
SELECT snapshot_id, change_type, COUNT(*) 
FROM ducklake_table_changes('my_ducklake', 'main', 'stats', 1, 2)
GROUP BY snapshot_id, change_type;
```

Or between snapshots 2 and 3:

```{sql}
#| connection: con
#| cache: false
SELECT snapshot_id, change_type, COUNT(*) 
FROM ducklake_table_changes('my_ducklake', 'main', 'stats', 2, 3)
GROUP BY snapshot_id, change_type;
```

Or we can count the number of insertions performed to create snapshot 1:

```{sql}
#| connection: con
#| cache: false
SELECT snapshot_id, COUNT(*) 
FROM ducklake_table_insertions('my_ducklake', 'main', 'stats', 1, 1)
GROUP BY snapshot_id;
```
:::

## Exploring gene expression results using SQL

### Tau-dependent gene expression changes across time 

Now that we have the results for multiple comparisons (aka contrasts) in place,
we can e.g. look for genes that are significantly up- or down-regulated
(FDR \< 5%) between Tau-transgenic and wildtype animals at *both* 6 and 8 
months of age.

To keep things readable, we create two Common Table Expressions (CTEs), each
extracting the results of one contrast (e.g. either `six_months` or 
`four_months`). Then we join the results and limit the output to 1000 random
genes that pass the FDR threshold. (We order the results by the `hash` of the
gene symbol to obtain a pseudo-random ordering.)

```{sql}
#| connection: con
#| cache: false
#| output.var: df
WITH six AS (
  SELECT symbol, logFC, "P.Value", "adj.P.Val"
  FROM stats
  WHERE study = 'Wang_2018' AND contrast = 'six_months' AND "adj.P.Val" < 0.05
),
eight AS (
  SELECT symbol, logFC, "P.Value", "adj.P.Val"
  FROM stats
  WHERE study = 'Wang_2018' AND contrast = 'eight_months' AND "adj.P.Val" < 0.05
)
SELECT s.symbol, 
       s.logFC six_logFC, s."P.Value" six_p,
       e.logFC eight_logFC, e."P.Value" eight_p
FROM six s
JOIN eight e ON s.symbol = e.symbol
ORDER BY hash(s.symbol, 1234)
LIMIT 1000;
```

We see that almost all genes that pass the FDR threshold at both time points are 
differentially regulated *in the same direction* at both time points (e.g. they
fall into the top right or bottom left quadrants). Their log fold changes are
highly correlated between the two ages as well (e.g. most of the points fall
close to the diagonal).

```{r}
#| fig-width: 5
#| fig-height: 5
df |>
  ggplot() +
  aes(x = six_logFC, y = eight_logFC) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) + 
  labs(x = "Six-months (log2 fold change)", 
       y = "Eight-months (log2 fold change)",
       title = "Tau vs WT",
       subtitle = "Significant differential expression at both ages.",
       caption = "FDR < 5%"
  )
```

## Examining the expression of individual genes using dplyr

We can also plot the normalized expression of a specific gene across all samples
collected in this study by querying the `exprs` table. For example, let's
examine the expression of the *Apoe* gene. 

Instead of writing the SQL query ourselves, we can use 
[dplyr](https://dplyr.tidyverse.org) 
to do the job for us. 

```{r}
df <- tbl(con, "exprs") |>
  dplyr::filter(study == "Wang_2018", symbol == "Apoe") |>
  dplyr::select("symbol", "cpm", "counts", "group")
dplyr::explain(df)  # <1> 
```
1. The `explain()` function shows us the query and the
plan that will be executed by `duckdb`

```{r}
df <- dplyr::collect(df)  # <2>
df |>
  DT::datatable(rownames = FALSE) |>
  DT::formatSignif(columns = "cpm", digits = 3)
```

2. Calling `collect()` triggers the actuall database scan and returns the
results in a `tibble`.

A scatter plot reveals that *Apoe* gene expression is stongly increased in the
Tau transgenic animals:

```{r}
#| fig-width: 5
#| fig-height: 3
data.frame(
  stringr::str_split_fixed(df$group, pattern = "_", 2)
) |>
  setNames(c("genotype", "age")) |>
  dplyr::mutate(genotype = factor(genotype, levels = c("WT", "rTg4510"))) |>
  dplyr::bind_cols(df) |>
  ggplot() +
  aes(x = age, y = cpm) +
  facet_wrap(~ genotype) +
  geom_point() +
  labs(x = "Age (months)", 
       y = "Normalized expression (CPM)",
       title = "Apoe"
  )
```

Right now, we only have data from a single study (albeit with four different
comparisons) in our `ducklake`. As we keep adding additional datasets and
contrasts, new parquet files will be added and our ability to compare and
contrast across studies will grow!

## Disconnecting our R session from the data lake

To disconnect, we first switch back to an in-memory `duckdb` connection, and
then detach from the data lake.

```{r}
#| cache: false
#| results: hide
dbExecute(conn = con, "USE memory; DETACH my_ducklake;")
dbDisconnect(con)
```

## Exploring gene expression results using python 🐍

Even though created and populated the `my_ducklake` database using the R API,
it is not tied to any specific language. That makes this setup
a great tool for teams that use multiple languages. 

### Connecting to our `ducklake`

For example, we can access the datalake from python using 
[duckdb's python API](https://duckdb.org/docs/stable/clients/python/overview) 
instead:

```{python}
#| eval: true
import duckdb

con = duckdb.connect(database = ":memory:")  # <1>
con.sql("INSTALL ducklake")  # <2>

con.sql(  # <3>
  f"""
  ATTACH 'ducklake:{r.db_file}' AS my_ducklake (READ_ONLY);
  """)
df = con.sql(
  """
  USE my_ducklake;
  SELECT * FROM stats LIMIT 1000;
  """
).to_df()  # <4>
```

1. First, we create an in-memory `duckdb` instance to communicate with the 
  metadata database. 
2. Next, we install the `ducklake` core extension, as we did in R above.
3. We attach our metadata database - in this example in read-only mode. 
  Here, we retrieve the path to the metadata database from the `db_file` R 
  variable we defined up (see above)[^4]. But we could just as well hard-code
  them or provide them in other ways.
4. Calling the `.to_df()` method at the end of the command returns the results
as a
[pandas DataFrame](https://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.html),
ready for downstream data wrangling or visualization.

[^4]: I love being able to use both R and python in the same quarto document,
and exchanging objects between them. Nicola Rennie has a very helpful [blog post](https://nrennie.rbind.io/blog/combining-r-and-python-with-reticulate-and-quarto/)
on how to get started.

Now that we have established the connection, we can run SQL queries from
python, e.g. retrieving 1000 rows from the (parquet-file backed) `stats` table 
[^5].

[^5]: Barry Smart demonstrates how to use python to manipulate a `ducklake`
in
[this blog post](https://endjin.com/blog/2025/06/ducklake-practice-hands-on-tutorial-core-features).

```{python}
#| eval: true
df.head()
```

### Differential expression visualized in a volcano plot

For example, we can create a volcano plot with the 
[matplotlib](https://matplotlib.org/) 
python module.

```{python}
#| fig-width: 5
#| fig-height: 5
import numpy as np
import matplotlib.pyplot as plt

df['neg_log10_pval'] = -np.log10(df['adj.P.Val'])
pval_thresh = 0.05
logfc_thresh = 1

plt.figure(figsize=(8,6))
plt.scatter(df['logFC'], df['neg_log10_pval'], color='grey', alpha=0.5)

sig_up = df[(df['logFC'] > logfc_thresh) & (df['adj.P.Val'] < pval_thresh)]
plt.scatter(sig_up['logFC'], sig_up['neg_log10_pval'], color='red', alpha=0.7,
            label='Upregulated')

sig_down = df[(df['logFC'] < -logfc_thresh) & (df['adj.P.Val'] < pval_thresh)]
plt.scatter(sig_down['logFC'], sig_down['neg_log10_pval'], color='blue', 
            alpha=0.7, label='Downregulated')
plt.axhline(-np.log10(pval_thresh), color='black', linestyle='dashed', 
            linewidth=1)

plt.axvline(logfc_thresh, color='black', linestyle='dashed', linewidth=1)
plt.axvline(-logfc_thresh, color='black', linestyle='dashed', linewidth=1)

plt.xlabel('log2 Fold Change')
plt.ylabel('-log10 Adjusted P-value')
plt.title('Volcano Plot')
plt.legend()
plt.show()
```

## Disconnecting our python session from the ducklake

As usual, we disconnect from the `ducklake` at the end of our analysis session.

```{python}
#| eval: true
con.close()
```

## Conclusions

In the past, I had considered storing gene expression data in parquet files, and
to manage their location, updates and versions myself. The `ducklake` extension
takes these tasks off my hand, and I am looking forward to exploring its 
capabilities further. 

I am already using a relational database to manage 
study, sample and gene metadata, e.g. a study's title, GEO accession, abstract,
etc or the properties of each of the samples that constitute an experiment. I
envision tracking additional information about comparisons in the same way 
(e.g. the specification of the linear model, or a human-readable description of
the contrast). 

The ability to join this metadata with the measurements (counts, CPMs) and the
statistical results (log fold changes, p-values, etc) could make `duckdb` my 
one-stop shop for data visualization and tertiary analysis. `ducklake` 
could also become the foundation for dashboards and other visualization tools,
allowing non-computational colleagues to explore a gene expression compendium
on their own.

I am a big fan of the well-designed classes offered by Bioconductor, and I take
advantage of great R packages every day. But I also appreciate the opportunity
to move results into a language-agnostic system, which can then fuel
new analyses across a multilingual data science organization.

## Reproducibility

::: {.callout-note collapse="true"}
### Session Information ℹ️
```{r}
sessioninfo::session_info()
```
:::
