---
title: "Non-compartmental PK analysis with the PKNCA R package - Part 2"
subtitle: "Key PK metrics"
author: "Thomas Sandmann"
date: "2025-05-04"
freeze: true
categories: [R]
editor:
  markdown:
    wrap: 72
format:
  html:
    toc: true
    toc-depth: 4
    code-tools:
      source: true
      toggle: false
      caption: none
editor_options: 
  chunk_output_type: console
---

## tl;dr

In a series of blog posts I will dive into non-compartmental analysis (NCA),
a set of tools from the field of Pharmacokinetics (PK). I am using the 
excellent 
[PKNCA R package](https://cran.r-project.org/package=PKNCA)
to estimate parameters, for both single- and multi-dose studies. This is a
new field for me, so buckle up and join me on my learning journey! This first
post focuses on the representation of PK data for use with `PKNCA`, and
demonstrates how top perform a basic analysis of a single-dose study.

## Example data: Theophylline treatment

The 
[PKNCA R package](https://cran.r-project.org/package=PKNCA),
written and maintained by 
[Bill Denney](https://orcid.org/0000-0002-5759-428X)
and Clare Buckeridge includes examples that use the `Theoph` dataset incluced
in base R. Let's start by exploring this dataset.

```{r}
suppressPackageStartupMessages({
  library("poorman")
  library("ggplot2")
  library("ggsci")
  library("patchwork")
  library("PKNCA")
  library("scales")
})
theme_set(theme_linedraw(14))
```


As in the my first post, I am using the `Theoph` dataset that is included with
R as an example. Today, I am focussing only on the `Subject`, `Time` and
`conc` columns.

```{r}
conc_df <- Theoph |>
  as.data.frame() |>
  poorman::select(Subject, Time, conc)
```

## Pharmacokinetic metrics

### $AUC_{last}$

The `PKNCA` R package supports the calculation of multiple estimates for the
"area under the curve", an estimate of total exposure to the drug, detailed in 
[this vignette](https://billdenney.github.io/pknca/articles/v05-auc-calculation-with-PKNCA.html).

For example, let's plot the measurements for Subject 1. At the end of the
study, e.g. after 25 hours, the `Theophylline` concentrations are still markedly
higher than at the start. 

```{r}
#| fig-width: 8
#| fig-height: 6
conc_subset <-  poorman::filter(conc_df$data, Subject == "1")
conc_subset |> 
  ggplot() + 
  aes(x = Time, y = conc) + 
  geom_point() + 
  geom_line(linewidth = 0.25) +
  scale_color_igv() +
  scale_x_continuous(limits = c(NA, 60), 
                     breaks = scales::pretty_breaks(n = 5)) + 
  labs(x = "Time (hr)", y = "Concentration (mg/L)",
       subtitle = "Subject 1",
       title = "Theophillin serum concentrations") +
  theme(panel.grid = element_blank())
```

The $AUC_{last}$ metric includes only the area under the curve below the points
that were _actually_ measured, e.g. in this example until 25 hours, and is 
calculated using the 
[trapezoid rule](https://en.wikipedia.org/wiki/Trapezoidal_rule)[^1]

[^1]: Danielle Navarro wrote a great blog entry on 
[Non-compartmental analysis](https://blog.djnavarro.net/posts/2023-04-26_non-compartmental-analysis/#a-simple-approach)
that illustrates (among other things) why the trapezoid rule is only a
simplified estimate of the AUC, and why it can over- and underestimate the
actual area.

```{r}
#| fig-width: 8
#| fig-height: 6
conc_subset <-  poorman::filter(conc_df$data, Subject == "1")
conc_subset |> 
  ggplot() + 
  aes(x = Time, y = conc) + 
  geom_ribbon(aes(ymin = min(conc), ymax = conc), fill="skyblue") +
  geom_point() + 
  geom_line(linewidth = 0.25) +
  scale_color_igv() +
  scale_x_continuous(limits = c(NA, 60), 
                     breaks = scales::pretty_breaks(n = 5)) + 
  labs(x = "Time (hr)", y = "Concentration (mg/L)",
       subtitle = bquote("Subject 1:" ~ AUC[last]),
       title = "Theophillin serum concentrations") +
  theme(panel.grid = element_blank())
```

Yet, there is clearly `Theophylline` left in the subject's serum at this time 
point, e.g. their overall drug dose is underestimated if we truncate the AUC 
at the last measured time point.

To estimate the _full_ area under the curve, beyond the last measured value,
we have to estimate how the drug concentration would continue to decline.
And to do _that_, we need to extrapolate beyond the last measured time point.
We will need to make some additional assumptions to do so.

### $AUC_{\inf}$

For many drugs, the decrease in concentration over time follows a first order
kinetics. If that's the case for `Theophylline`, then we should 
observe a linear log(concentration) ~ Time relationship - at least at the end
of the time course, when the majority of the orally available drug has
already entered the serum.

After log transforming the y-axis, we observe just that:

```{r}
conc_min <- min(conc_subset$conc)
conc_subset |>
  ggplot() + 
  aes(x = Time, y = conc) + 
  geom_ribbon(aes(ymin = conc_min, ymax = conc), fill="skyblue") +
  geom_point(size = 3) + 
  geom_line(linewidth = 0.25) +
  scale_color_igv() +
  scale_y_continuous(transform = scales::log10_trans(), 
                     limits = c(conc_min, NA)) + 
  scale_x_continuous(limits = c(0, 60), 
                     breaks = scales::pretty_breaks(n = 5)) + 
  annotation_logticks(sides = "l") + 
  labs(x = "Time (hr)", y = "Concentration (mg/L)",
       subtitle = bquote("Subject 1:" ~ AUC[last]),
       title = "Theophillin serum concentrations") +
  theme(panel.grid = element_blank())
```

Next, we fit a line through the last four points[^2], regressing 
`log10(concentration)` on `Time` as the only predictor[^3].

[^2]: Which / how many points to include in the linear fit is another choice the
analyst makes. The minimum number of points and the choice between different
potential linear fits is governed by several of the 
[PKNCA.options](https://billdenney.github.io/pknca/articles/v40-options-for-controlling-PKNCA.html)
that make the business rules that are applied transparent.

[^3]:Please note that we are implementing this procedure for illustration
only; the `PKNCA` package includes helper functions to interpolate or
extrapolate measurements. Please consult the
[help pages](https://billdenney.github.io/pknca/reference/interp.extrap.conc.html)
for details.

```{r}
fit <- lm(log10(conc) ~ Time, data = tail(conc_subset, n = 4))  # <1>
summary(fit)  # <2>
```

1. Because our concentration data.frame is sorted by increasing Time, we 
can simply subset it to the last 4 rows.
2. The fit explains a large fraction of the variance for these (last 4)
measurements, as evidenced by an R-squared value close to 1.

Next, we can use the linear model to extrapolate beyond the last measured
data point, and also predict when the concentration will drop to zero - or the
lower limit of quantitation (`LLQ`). We don't know the `LLQ` for the assay
used in the `Theoph` dataset, we will simply use the lowest reported
concentration ($C_{min}$, observed at time zero).

```{r}
time_min = (
  log10(conc_min) - fit$coefficients["(Intercept)"]
  ) / fit$coefficients["Time"]
```

Based on our model, `Theophylline` should have returned to baseline levels
after `r unname(ceiling(time_min))` hours.

We can also predict the remaining `Theopylline` levels at specific times. Let's 
pick 4 evenly-spaced time points between the last measured one 
(`r max(fit$model$Time)` hours) and `r unname(ceiling(time_min))` hours and
use the `predict.lm` function to obtain the prediced concentrations.

```{r}
extra_time <- seq(max(fit$model$Time), time_min, length.out = 4)
extrapolated <- data.frame(
  conc = 10^predict.lm(fit, newdata = data.frame(Time = extra_time)),
  Time = extra_time,
  origin = "predicted"
)
extrapolated
```

We can add these extrapolated measurements to our dataset and plot the
both the observed and predicted measurements together - first with a 
log10 transformed y-axis, and then the untransformed concentrations:

```{r}
#| fig-width: 12
#| fig-height: 5
df_combined <- conc_subset |> 
  poorman::select(Time, conc) |>
  poorman::mutate(origin = "observed") |>
  poorman::bind_rows(extrapolated)

p1 <- df_combined |>
  ggplot() + 
  aes(x = Time, y = conc) + 
  geom_ribbon(
    aes(ymin = conc_min, ymax = conc), fill="skyblue",
    data = ~ filter(.x, origin == "observed")) +
  geom_ribbon(
    aes(ymin = conc_min, ymax = conc), fill="pink",
    data = ~ filter(.x, origin == "predicted")) +
  scale_color_manual(values = c(observed = "blue", predicted = "red"), 
                     name = element_blank()) + 
  scale_fill_discrete(name = element_blank()) + 
  geom_point(size = 3, aes(color = origin)) + 
  geom_line(linewidth = 1, aes(color = origin)) +
  geom_line(data = extrapolated, color = "red", linewidth = 1) +
  scale_y_continuous(transform = scales::log10_trans(), 
                     limits = c(conc_min, NA)) + 
  scale_x_continuous(limits = c(0, 60), 
                     breaks = scales::pretty_breaks(n = 5)) + 
  annotation_logticks(sides = "l") + 
  labs(x = "Time (hr)", y = "Concentration (mg/L)",
       subtitle = bquote("Subject 1:" ~ AUC[inf]),
       title = "Log linear concentration ~ time curve") +
  theme(panel.grid = element_blank())

p2 <- df_combined |>
  ggplot() + 
  aes(x = Time, y = conc) + 
  geom_ribbon(
    aes(ymin = conc_min, ymax = conc), fill="skyblue",
    data = ~ filter(.x, origin == "observed")) +
  geom_ribbon(
    aes(ymin = conc_min, ymax = conc), fill="pink",
    data = ~ filter(.x, origin == "predicted")) +
  scale_color_manual(values = c(observed = "blue", predicted = "red"), 
                     name = element_blank()) + 
  scale_fill_discrete(name = element_blank()) + 
  geom_point(size = 3, aes(color = origin)) + 
  geom_line(linewidth = 1, aes(color = origin)) +
  geom_line(data = extrapolated, color = "red", linewidth = 1) +
  scale_y_continuous(limits = c(conc_min, NA)) + 
  scale_x_continuous(limits = c(0, 60), 
                     breaks = scales::pretty_breaks(n = 5)) + 
  labs(x = "Time (hr)", y = "Concentration (mg/L)",
       subtitle = bquote("Subject 1:" ~ AUC[inf]),
       caption = "Red: log-linear extrapolation",
       title = "Linear concentration ~ time curve") +
  theme(panel.grid = element_blank())
p1 + p2 + plot_layout(guides = 'collect')
```

The linear curve shows the gradual decline in `Theophylline` concentrations,
and we can use this curve to estimate the complete $AUC_{\inf}$ metric[^4].

[^4]: Actually, this is not the full story. the `PKNCA` package can calculate
different $AUC_{\inf}$ estimates: $AUC_{\inf,obs}$ and $AUC_{\inf,pred}$,
depending on whether the observed or predicted concentrations for the last
time point above the LLQ (e.g. $t_{last}$) are used. On top of that, there is
also the $AUC_{all}$ version, which is useful when you have later observations
that fell below the LLQ. Please see the
[vignette](https://billdenney.github.io/pknca/articles/v05-auc-calculation-with-PKNCA.html)
for more details - and consult your friendly neighborhood PK/PD expert for
advice on which metrics are used in your organization.


## Reproducibility

<details>
<summary>
Session Information
</summary>

```{r}
sessionInfo()
```

</details>

