---
title: "Non-compartmental PK analysis with the PKNCA R package"
subtitle: "First step into the pharmacokinetics world"
author: "Thomas Sandmann"
date: "2025-04-27"
freeze: true
categories: [R]
editor:
  markdown:
    wrap: 72
format:
  html:
    toc: true
    toc-depth: 4
    code-tools:
      source: true
      toggle: false
      caption: none
editor_options: 
  chunk_output_type: console
---

## tl;dr

Pharmacokinetics (PK) studies how drugs are absorbed, distributed, metabolized
or excreted (ADME) over time. Analyses usually start with plotting the
measured drug concentrations versus time, followed by the calculation of one or
more variables, e.g. estimates of drug _exposure_ over time such as 
minimum or maximum drug concentratiosn ($C_{min}$ and $C_{max}$), 
the time at which $C_{max}$ was reached ($t_{max}$), the total exposure - often
based on the area under curve (AUC), etc.

To describe the behavior of a drug, empirical or mechanistic models are used, 
and the latter can be sub-grouped into `compartmental` or `non-compartmental`
approaches [^1]. Here, I am focusing exclusively on non-compartmental analysis
(NCA), which is widely used and offer standard algebraic solutions for many
types of questions [^2].


[^1]: [Applied Biopharmaceutics and Pharmacokinetics](https://accesspharmacy.mhmedical.com/book.aspx?bookid=1592)
by Shargel and Yu discusses the distinction between different pharmacokinetic
approaches and their pros and cons in detail.

[^2]: Compartmental models usually rely on statistical models represented as
partial differential equations. The model is fit to the data using e.g. least
squares regression, maximum-likelihood or Bayesian techniques and often require
numerical solutions to non-linear regression problems. They also require the
analyst to make explicit and implicit assumptions, e.g. about the the
[reaction order](https://en.wikipedia.org/wiki/Rate_equation)
governing the rates at which drug concentration changeover time.

## Non-compartmental analysis

NCA is based on estimating PK parameters directly from the concentration-time
curve, without requiring a mathematical model that describes the entire behavior
of the drug [^3]. It is appropriate when 

1. Exposure increases proportionally with drug dose and the PK parameters don't
   change over time (e.g. there aren't any saturation effects, etc)
2. The drug is eliminated from the same pool that the concentration measurements
   are obtained from.
3. There are not other sources of the drug than those in the measured pool.
4. Sufficient data is available, e.g. 12 or more concentation time points for
   a single-dose administration.

[Source: Shargel an Yu, 7th edition, Chapter 25](https://accesspharmacy.mhmedical.com/book.aspx?bookid=1592)

[^3]: NCA approaches are sometimes referred to as "model-free", but that's not
entirely correct. For example, to calculate the full exposure of a subject it
is often necessary to extrapolate beyond the final measurement. Typically, a
first-order elimination process is assumed to estimate the residiual drug
concentrations, e.g. expecting a log-linear decrees toward zero, to calculate
the $AUC_\infty$ value.

## Software solutions

There are many commercial tools to perform PK analyses, many of which can
perform both NCA and compartmental analyses. The relevant
[CRAN Task View](https://cran.r-project.org/web/views/Pharmacokinetics.html)
also list multiple options, including long-standing solutions such as 
[ncappc](https://cran.r-project.org/package=ncappc), 
published in 2015, or
[NonCompart](https://cran.r-project.org/package=NonCompart),
published in 2016.

Another solution, the
[PKNCA R package](https://cran.r-project.org/package=PKNCA), 
also published in 2015, stood out to me because of the depth and breadth of its 
[documentation](https://billdenney.github.io/pknca/), 
including multiple vignettes. In addition, the `PKNCA` R package also allows
users to specify business rules that impact the final results, e.g. how to 
deal with missing values - either within a set of measurements or at the end
of a time series.

## Theophillin treatment data

The 
[PKNCA R package](https://cran.r-project.org/package=PKNCA),
written and maintained by 
[Bill Denney](https://orcid.org/0000-0002-5759-428X)
and Clare Buckeridge includes examples that use the `Theoph` dataset incluced
in base R. Let's start by exploring this dataset.

```{r}
suppressPackageStartupMessages({
  library("poorman")
  library("ggplot2")
  library("ggsci")
  library("PKNCA")
  library("scales")
})
theme_set(theme_linedraw(14))
```

The `datasets::Theoph` dataset contains `r nrow(Theoph)` measurements from
`r nlevels(Theoph$Subject)` subjects, each treated with a single oral dose of
[Theophylline](https://en.wikipedia.org/wiki/Theophylline), 
a drug classically used to treat e.g. Asthma or chronic obstructive pulmonary
disease (COPD) [^4].

[^4]: Theophillin is found naturally e.g. in tea; as a drug it has mainly been
replaced by newer pharmaceuticals - see e.g. [this review](https://pmc.ncbi.nlm.nih.gov/articles/PMC11170423/)
for more details.

```{r}
summary(datasets::Theoph)
```

At each time point, the concentration of `Theophillin` in the serum was
recorded. Most of the `r length(unique(Theoph$Dose))` different oral doses 
levels were only assessed in a  single subject, but two subjects received 
`4.4 mg/kg` or `4.53 mg/kg` of the drug.

```{r}
#| fig-width: 7.5
#| fig-height: 5
label_function <- function(x) paste(x, "mg/kg")

Theoph |>
  poorman::mutate(
    Subject = factor(Subject, levels = unique(Subject)),
    Dose = factor(Dose, levels = sort(unique(Dose)))) |>
  ggplot() +
  aes(x = Time, y = conc, color = Subject) + 
  geom_point(size = 1) +
  geom_line(linewidth = 0.25) +
  scale_color_igv() +
  labs(x = "Time (hr)", y = "Concentration (mg/L)",
       title = "Theophillin serum concentrations by Dose") +
  facet_wrap(~ Dose, labeller = labeller(Dose = label_function)) +
  theme(panel.grid = element_blank())
```

Because the drug was given orally and needed to be absorbed through the
gut before it reaches the serum (and circulates throughout the body), its
maximum concentration is reached after several ours, and it declines slowly over
the course of roughly a day.

For analysis with the `PKNCA` package, we need to represent the data in (at 
least) data.frames, one for `dose` information and the other for `concentration`
measurements. (A third data.frame can be used to specify `intervals` of 
interest, see below.)

Let's start by capturing the concentration data by passing the relevant columns
of the `Theophil` dataframe to the `PKNA::PKNCAconc` function.

```{r}
conc_df <- PKNCAconc(data = Theoph, formula = conc ~ Time|Subject)  # <1>
conc_df  # <2>
```

1. The `Subject` column is encoded as a factor. The right hand side of the
  formula groups the data by subject, reflecting the multi-subject design of
  this dataset. (The right-most variable is always assumed to correspond to an
  individual.)
2. The `conc_df` object is an instance of the `PKNCAconc` class. 

Next, we subset the `Theoph` dataset to the first time point for each subject,
e.g. `Time == 0`, when they received the single drug dose and pass these rows
into the `PKNCAdose` constructor.

```{r}
dose_df <- PKNCAdose(data = subset(Theoph, Time == 0),
                     formula = conc ~ Time|Subject)
dim(dose_df$data)  # <3>
```

3. The `dose_df` object is an instance of the `PKNCAdose` class, and it`s 
  `data` element contains a single row for each subject (e.g. their single 
  dose).

::: {.callout-note collapse="true"}
### Sparse data

Sometimes, datasets are *sparse*, e.g. no subject / animal has data for the 
entire time course. (An extreme example are endpoints that can only be
measured post mortem; for pre-clinical _in vivo_ studies, animals might be
sacrificed at different times and each contribute only a single data point
to the PK dataset.)

To perform NCA on sparse data, simply add the `sparse=TRUE` argument to the 
`PKNCAdose` call.

:::

Next, we combine both objects by passing them the `PKNCAdata` function.

```{r}
pk_data <- PKNCAdata(conc_df, dose_df)  # <4>
results <- pk.nca(pk_data)  # <5>
summary(results)  # <6>
```

4. Because we omit the optional `intervals` argument to the `PKNCAdata`
  function, default intervals are defined for us automatically, and a minimal
  set of parameters to estimate is included. (More on that below.)
5. The `pk.nca` data performs the actual calculations, separately for each
  interval.
6. Finally, the `summary` function returns the results & confidence intervals 
  in the form of a data.frame - with one for each interval and group.

We obtain summarized results across all subject for the following
parameters[^5]:

- `N` the number of subjects
- `auclast`: $AUC_{last}$, e.g. the area under the concentration time curve from
  the beginning of the interval to the last concentration above the limit of
  quantification, represented as geometric mean and geometric coefficient of 
  variation.
- `cmax`: $C_{max}$, e.g. the maximum observed concentration
- `tmax`: $t_{max}$, e.g. the time of the maximum observed concentration,
  represented as median & range
- `half.life`: $t_{\frac{1}{2}}$, e.g. the (terminal) half-life, represented
  as arithmetic mean and standard deviation
- `aucinf.obs`: $AUC_{obs}^{\infty}$ e.g. the area under the concentration
  time curve from the beginning of the interval to infinity with extrapolation 
  to infinity from the observed last measured concentration ($C_{last}$), 
  represented as geometric mean and geometric coefficient of variation.

The summary has two rows, one for each of the two default intervals:

1. 0-24 hours
2. 0-Infinity

and parameters are estimated when their requirements are met, e.g. $AUC_{last}$
can be calculated based on the first interval, but $AUC_{obs}^{\infty}$
requires measurements taken at the last timepoint (25 hours).

[^5]: A full list of available parameter is 
[available here](https://billdenney.github.io/pknca/articles/v03-selection-of-calculation-intervals.html#parameters-available-for-calculation-in-an-interval)

::: {.callout-note collapse="false"}
### Redefining aggregation functions

The `PKNCA` package includes predefined aggregation statistics for each of
the available parameters, but it also allows us to overwrite them.

For example, by default the _half-life_ is summarized by calculating the
arithmetic meand and the standard deviation. If we prefer e.g. calculating
the median and median absolute deviation (MAD) instead, we can define these 
new business rules with the `PKNCA.set.summary`, along with a human-readable
description:

```{r}
PKNCA.set.summary(
  name="half.life",
  description="median and median absolute deviation",
  point=\(x) median(x, na.rm = TRUE),
  spread=\(x) mad(x, na.rm = TRUE),
  rounding=list(signif = 3)
)
```

**Note:** Please take a look at the predefined aggregation functions, e.g. 
`business.mean` or `business.sd` to understand how they behave when there are
any / many missing values. You might want to implement a similar behavior in
your own summarization functions.

:::

### Accessing estimates for individual subjects

The `summary` function automatically summarizes results, using an appropriate
method (e.g. arithmetic mean, median, etc). If no grouping has been defined (as
in the case shown above), summary statistics are aggregated across the entire
dataset. 

To access the results _before_ aggregation, e.g. for each individual subject,
we can access the `result` slot of the `pk.nca()` output. (For convenience,
that authors defined a `as.data.frame()` S3 method that extracts this slot for
us.)

For each subject and interval (identified by the `start` and `end` columns), 
we see the estimate (`PPORRES`) for every parameter (`PPTESTCD`). 

```{r}
as.data.frame(results) |>
  head(n = 14)
```

The object not only contains the parameters that were reported in the summary,
but also those that were computed because they were required to estimate the
requested ones.

```{r}
#| warning: false
results |>
  as.data.frame() |>
  poorman::distinct(PPTESTCD)
```

## Defining intervals

So far, we have relied on the `PKNCAdata` function's ability to define default
intervals. 

As we observed above, one of the default intervals spans the entire study, e.g.
unless instructed otherwise, the `PKNCAdata` function will define an interval
from `start=0` to `end=Inf` by default.

The `Theophil` dataset's last time point was collected after 25 
hours. To demonstrate the flexibility of the `PKNCAdata` function, let's define
return _different_ parameters for two separate intervals:

An interval object is a data.frame with both `start` and `end` columns, as well
as columns for each of the parameters we would like to estimate [^6]. For 
example,  

1. 0 to 24 hours: `auclast`, `r.squared` and `adj.r.squared` parameters
2. 0 to 25 hours: `auclast`, `lambda.z` and `tlast` parameters

**Note:** Overlapping intervals, e.g. as in this example, are no problem.

[^6]: The full list of supported parameters is available 
[here](https://billdenney.github.io/pknca/articles/v03-selection-of-calculation-intervals.html#parameters-available-for-calculation-in-an-interval).

```{r}
interval_df <- data.frame(
  start = 0,  # <1>
  end = c(24, 25),  # <2>
  auclast = TRUE,
  r.squared = c(TRUE, FALSE),
  adj.r.squared = c(TRUE, FALSE),
  lambda.z = c(FALSE, TRUE),
  tlase = c(FALSE, TRUE)
)
interval_df
```

1. To create a data.frame with two rows, we can rely on R's vector recycling
  behavior to fill in the required number of identical entries.
2. When values differ between rows, we need to explicitly list each value.

```{r}
pk_data <- PKNCAdata(conc_df, dose_df, intervals = interval_df)  # <3>
results <- pk.nca(pk_data)
summary(results)
```

3. The data.frame is passed to the `PKNCAdata` function via the `intervals`
argument.

**Note:** Although we only specified the parameters we were interested in,
the `PKNCAdata` function automatically also requested all other parameters that
were needed for these calculations for us. As before, they are available in the
`results` object, but only those we explicitely requested are aggregated and 
included in the summary.

## Nested groups

The `Theoph` dataset contains one more column that we haven't used, yet, the
body weight of each subject. Let's split this statistic at the median, to create
two subgroups of subjects: `A` and `B`.

```{r}
Theoph$Group <- factor(ifelse(Theoph$Wt > median(Theoph$Wt), "A", "B"))
```

How can we obtain parameter estimates _separately_ for each of these subgroups?
Let's return to the beginning of our analysis and define the concentration
and dose data.frames, but also include the `subgroup` column this time.

The rest of our code remains unchanged, as the `summary` function will
aggregate the estimates at the levels of the `Group` variable automatically.

```{r}
conc_df <- PKNCAconc(data = Theoph, 
                     formula = conc ~ Time|Group + Subject)  # <1>
dose_df <- PKNCAdose(data = subset(Theoph, Time == 0),
                     formula = conc ~ Time|Group + Subject)
pk_data <- PKNCAdata(conc_df, dose_df, intervals = interval_df)  # <2>
results <- pk.nca(pk_data)  # <3>
summary(results)  # <4>
```

1. We include the `Group` covariate in the formula, to the _left_ of the
  `Subject` variable - because the final variable is interpreted as the 
  individual subject by default. (This can be overriden by specifying the 
  `subject` argument.)
2. We combine the three data.frames (including the custom intervals we 
  defined above).
3. The `pk.nca` estimates the subject-level results
4. We obtain the summary statistics - aggregated at the `Group` level.

::: {.callout-note collapse="false"}
### Varying intervals by group

What if we wanted to request different parameters for group `A` than for group
`B`? Easy - we simply include a `Group` column in our `intervals_df` data.frame.
The `PKNAdata` function automatically joins data.frames based on column names.

For example, let's request `r.squared` for group `A`, and `adj.r.squared` for
group `B`:

```{r}
interval_df <- data.frame(
  start = 0,  # <1>
  end = Inf,
  Group = c("A", "B"),  # <2>
  r.squared = c(TRUE, FALSE),  # <3>
  adj.r.squared = c(FALSE, TRUE)
)
interval_df
```

1. We only focus on a single interval here, spanning the entire experiment.
2. We add a new column to the data.frame, with one row for each of the groups.
3. We can toggle the parameter request for each group by setting the boolean
  values for each row.

The rest of our code remains unchanged - the interval definitions do all of the
heavy lifting!

```{r}
conc_df <- PKNCAconc(data = Theoph, 
                     formula = conc ~ Time|Group + Subject)
dose_df <- PKNCAdose(data = subset(Theoph, Time == 0),
                     formula = conc ~ Time|Group + Subject)
pk_data <- PKNCAdata(conc_df, dose_df, intervals = interval_df)
results <- pk.nca(pk_data)
summary(results)
```

The summary now return two rows, one for each `Group` and the same 0 -> Inf
interval.

**Note:** We could also have requested multiple intervals at the same time by
specifying additional rows in the data.frame, as in the examples above.

:::

## Multi-dose studies

In the `Theoph` study every subject received only a single dose of the drug.
To demonstrate that the `PKNCA` R package can handle multi-dose studies as well,
we will simulate a longer time series with multiple doses using
[superposition](https://www.certara.com/knowledge-base/the-superposition-principle/).

### Superposition

We follow the example from the
[PKNCA vignette](https://billdenney.github.io/pknca/articles/Superposition.html)
and simulate the profile we should expect if all subjects had received
7 doses every 12 hours over a 96 hour interval.

::: {.callout-note collapse="true"}
### Assumptions

The principle of superposition is based on the assumptions that 

1. The drug is eliminated by first-order kinetics.
2. Subsequent doses are not influenced by previous ones, e.g. there is no change
in physiology, drug metabolism is not altered due to enzyme saturation or 
induction, etc.

:::

```{r}
Theoph |>
  as.data.frame() |>  # <1>
  poorman::filter(Time == 0)
```

1. The `Theoph` dataset is not a regular data.frame, but a `groupedData` object
(defined in the `nlme` R package). We coerce it to a regular data.frame to use
`poorman` (or `dplyr`) functions to subset and modify it.

You might have noticed that the concentration at the first time point (0 hours)
is not exactly zero. This could be due to measurement error (e.g. the lower
limit of quantitation of the assay), or because the subject's serum contained
theophylline from another source[^7].

[^7]: Trace amounts of theophylline are naturally present in tea, coffee, 
chocolate, yerba maté, guarana, and kola nut (source: [Wikipedia](https://en.wikipedia.org/wiki/Theophylline)).

To extrapolate the data via superposition, though, we need to include a
baseline value of zero, so (**just for didactic purposes**), we reset that
value manually:

```{r}
theo_corrected <- Theoph |>
  as.data.frame() |>
  poorman::mutate(conc = ifelse(Time == 0, 0, conc))  # <2>
theo_corrected |>
  poorman::filter(Time == 0)
```

2. We set the concentration measurement for the first time point to zero.

Now, we can superimpose the profile, assuming that a new dose of the drug
is provided every 12 hours.

```{r}
pk_multi <- theo_corrected |>
  PKNCAconc(formula = conc ~ Time|Subject) |>
  PKNCA::superposition(
    tau = 4 * 24,  # <3> 
    n.tau = 1,  # <4>
    dose.times=seq(0, 3 * 24, by = 12)  # <5>
  )
```

3. The total dosing interval (from first dose to last observation).
4. The number of `tau` dosing intervals to simulate, set to one because we
  want to simulate a single interval starting with a concentration of zero.
5. Specific dosing times within the interval.

```{r}
#| fig-width: 8
#| fig-height: 5
pk_multi |>
  poorman::mutate(
    Subject = factor(Subject, levels = unique(Subject))
  ) |>
  ggplot() +
  aes(x = time, y = conc, color = Subject) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.25) +
  scale_color_igv() +
  labs(x = "Time (hr)", y = "Concentration (mg/L)",
       title = "Multi-dose study (superposition)") +
  theme(panel.grid = element_blank())
```

## Excluding results

[Excluding results](https://billdenney.github.io/pknca/articles/v07-post-processing.html#modifying-results)

## Reproducibility

<details>
<summary>
Session Information
</summary>

```{r}
sessionInfo()
```

</details>

